---
title: "Maxent SDM using BioClim and Land Use data"
author: "Pablo Deschepper"
date: "`r Sys.Date()`"
output: html_document
params:
  Species:
    label: "Species:"
    value: "Vipera latastei"
    input: text
  Database:
    label: "Database:"
    value: inaturalist
    input: select
    choices: [gbif, inaturalist, observation]
  Limit:
    label: "Number of records:"
    value: "1000"
    input: numeric
  Workdir: 
    label: "Select your working directory:"
    value: "C:/Documents and Settings/pdeschepper/Desktop/Maxent"
    input: text
  Country:
    label: "2 letter country code:"
    value: ""
    input: text
  Continent:
    label: "Continent"
    value: ""
    input: select
    choices: [all, africa, antarctica, asia, europe, north_america, oceania, south_america]
    
  
---
```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(warning = FALSE, message = TRUE)
knitr::opts_knit$set(root.dir = params$Workdir)

```

- **Species = *`r params$Species`* **
- Using records from followning database: `r params$Database`
- The number of records is limited to `r params$Limit` records

The working directory = *`r params$Workdir`*<br>
note: The script will download all BioClim and Land Use data upon first use of the script, translating into a longer processing times. The script will run faster after first use.

```{r libraries, include=FALSE}
library(dismo)
library(raster)
library(rgbif)
library(sf)
library(rmaxent)
library(rJava)
library(rasterVis)
library(viridis)
library(geodata)
library(ggplot2)
library(ggmap2)
library(stringr)
library(rnaturalearth)
library(caret)

```

```{r Get species records, include=FALSE, echo=FALSE}

#### Search for species occurrences in GBIF ####
# this option is straighforward but does not allow a lot of options
ContinentParam <- c()
if (params$Continent == "all") {
  ContinentParam <- ""} else {ContinentParam <- params$Continent}

occurrences <- rgbif::occ_data(scientificName = params$Species,
                              limit = params$Limit,
                              hasCoordinate = TRUE,
                              hasGeospatialIssue = FALSE,
                              country = params$Country,
                              continent = ContinentParam)
if (params$Database == "inaturalist") {
  occurrences <- data.frame(occurrences$data[, c("decimalLongitude", "decimalLatitude", "occurrenceID")])
  occurrences <- occurrences[str_detect(occurrences$occurrenceID, "inaturalist"),] %>% na.omit()
  recs <- nrow(occurrences)
  print(paste("Using",recs,"inaturalist records"))
  } else if (params$Database == "observation") {
    occurrences <- data.frame(occurrences$data[, c("decimalLongitude", "decimalLatitude", "occurrenceID")])
    occurrences <- occurrences[str_detect(occurrences$occurrenceID, "observation"),] %>% na.omit()
    recs <- nrow(occurrences)
    print(paste("Using",recs,"observation.org records"))
  } else {
    occurrences <- data.frame(occurrences$data[, c("decimalLongitude", "decimalLatitude", "occurrenceID")]) %>% na.omit()
    recs <- nrow(occurrences)
    print(paste("Using",recs,"gbif records"))
  }
  
if (recs < 100) {
  stop("Fewer than 100 species records: Process aborted.")
}

# Create a SpatialPointsDataFrame object for the occurrence data
spdf <- SpatialPointsDataFrame(coords = occurrences[, c("decimalLongitude", "decimalLatitude")], data = occurrences)
# Bounding box of distribution
bbox <- st_bbox(spdf) 
points <- data.frame(occurrences[, c("decimalLongitude", "decimalLatitude")])
```

```{r (Down)Load BioClim and Land Use data, include=TRUE, echo=FALSE}

#### Load BioClimatic data ####
# Create an extent object for the study area using bounding box coordinates
# By adding 2, we add 1 degree at each side
study_area_ext <- extent(bbox) + 2
# Create a raster object for the study area
study_area_ras <- raster::raster(study_area_ext)
# Set the resolution of the raster
res(study_area_ras) <- 10
# Download the WorldClim data for the study area and do some plotting
# Check if the path exists
if (!dir.exists("worldclim_data")) {
  # Download WorldClim data if the directory does not exist
  worldclim_data <- geodata::worldclim_global(var = 'bio', res = 10, download = TRUE, path = "worldclim_data")
  print("BioClim data has been downloaded and saved to the working directory.")
} else {
  print("BioClim data already exists and is loaded from the working directory.")
  rast_list <- c(list.files("worldclim_data/climate/wc2.1_10m/", full.names = TRUE)) %>%
    .[grepl("bio_([1-9]|1[0-9])", .)]
  worldclim_data <- terra::rast(rast_list)
  }
# convert from SpatRaster to Raster object while keeping all layers
worldclim_data_raster <- as(worldclim_data, "Raster")
# Crop the WorldClim data to the study area
worldclim_data_crop <- crop(worldclim_data_raster, study_area_ras)
# (optional) plot occurrences 
data(outline, package = "ggmap2")

#Download BioClim data for two future time periods
fut <- geodata::cmip6_world(model='ACCESS-ESM1-5', ssp='245', time='2041-2060', var='bioc', download=F, res=10, path='worldclim_data')
fut2 <- geodata::cmip6_world(model='ACCESS-ESM1-5', ssp='245', time='2061-2080', var='bioc', download=F, res=10, path='worldclim_data')
fut_crop <- crop(fut, study_area_ras)
fut2_crop <- crop(fut2, study_area_ras)
names(fut_crop) <- names(worldclim_data_crop)
names(fut2_crop) <- names(worldclim_data_crop)
```

```{r Load Land Use data , include=FALSE}

### load land use data ####
elevation <- geodata::elevation_global(bbox, path = "Elevation", res = 10)
footprint <- geodata::footprint(year = 2009, path = "Footprint")
population <- geodata::population(year = 2020, path = "Population", res = 10)
trees <- geodata::landcover(var = "trees", path = "Trees")
grassland <- geodata::landcover(var = "grassland", path = "Grassland")
shrubs <- geodata::landcover(var = "shrubs", path = "Shrubs")
built <- geodata::landcover(var = "built", path = "Built")
bare <- geodata::landcover(var = "bare", path = "Bare")
wetland <- geodata::landcover(var = "wetland", path = "Wetland")
water <- geodata::landcover(var = "water", path = "Water")


# make everything 10-min spatial resolution (not the case for land cover and footprints which are in 30-second res)
# and crop to study area
rescale_and_crop <- function(raster) {
  cropped <- crop(raster, study_area_ras)
  rescaled <- terra::aggregate(cropped, fact = 20, fun = 'mean')
  rasterized <- raster::raster(rescaled)
  # resampled <- terra::resample(rasterized, study_area_ras)
  return(rasterized)
}
trees_rc <- rescale_and_crop(trees)
grassland_rc <- rescale_and_crop(grassland)
shrubs_rc <- rescale_and_crop(shrubs)
built_rc <- rescale_and_crop(built)
bare_rc <- rescale_and_crop(bare)
water_rc <- rescale_and_crop(water)
wetland_rc <- rescale_and_crop(wetland)
footprint_rc <- rescale_and_crop(footprint)
elevation_rc <- crop(elevation, study_area_ras)
population_rc <- crop(population, study_area_ras)


# combine all of the landuse rasters into a rasterbrick
landuse <- c(trees_rc, grassland_rc, shrubs_rc, built_rc, bare_rc, footprint_rc,
             elevation_rc, population_rc, water_rc, wetland_rc) 
landuse_rasterbrick <- do.call(stack, landuse)

```



## Species records
This plot shows the georeferenced species records that will be used to train the Species Distribution Models (SDM) using MaxEnt. It is good to maximize the number of records.

**`r recs` records** of `r params$Species` are used for processing.

```{r Records plot, include=TRUE, echo=FALSE}
ggplot() + geom_point(data = points, aes(x=decimalLongitude, y = decimalLatitude), col="red") +
  geom_sf(data=outline, fill = "transparent", color="black") +
  coord_sf(xlim = c(bbox$xmin, bbox$xmax), ylim = c(bbox$ymin, bbox$ymax), expand = F) + theme_bw()
```


```{r Combine BoiClim and Land use data into predictors, include=FALSE}
### Make predictors dataset ####
# aggregate bioclimatic and land use date in one dataset
# align the extents
worldclim_data_aligned <- crop(worldclim_data_crop, extent(landuse_rasterbrick))
# resample to match resolution
if(!all(res(worldclim_data_aligned) == res(landuse_rasterbrick))) {
  worldclim_data_aligned <- resample(worldclim_data_aligned, landuse_rasterbrick, method = 'bilinear')
}
predictors <- stack(c(worldclim_data_aligned, landuse_rasterbrick))

```

```{r Clean the Full, BioClim and Land Use dataset by removing correlation, include=FALSE}
# Clean dataset for correlation > 0.75
datasets <- list(predictors = predictors, 
                 worldclim_data_aligned = worldclim_data_aligned, 
                 landuse_rasterbrick = landuse_rasterbrick)
cleaned_datasets <- list()
# Loop through each dataset and remove highly correlated variables
for (dataset_name in names(datasets)) {
    dataset_df <- as.data.frame(datasets[[dataset_name]], na.rm = TRUE)
    cor_mat <- cor(dataset_df, method = 'spearman', use = 'complete.obs')
    highly_correlated <- findCorrelation(cor_mat, cutoff = 0.75)
    # Only remove layers if there are correlated layers to remove
    if (length(highly_correlated) > 0) {
      cleaned_dataset <- datasets[[dataset_name]][[ -highly_correlated ]]
    } else {
      cleaned_dataset <- datasets[[dataset_name]]
    }
    cleaned_datasets[[dataset_name]] <- cleaned_dataset
    print(paste("Cleaned layers in", dataset_name, ":"))
    print(names(cleaned_dataset))
}

# Access the cleaned datasets
predictors_cleaned <- cleaned_datasets$predictors
worldclim_data_aligned_cleaned <- cleaned_datasets$worldclim_data_aligned
landuse_rasterbrick_cleaned <- cleaned_datasets$landuse_rasterbrick
# Also subset (clean) the future BioClim data
fut_crop_cleaned <- fut_crop[[names(worldclim_data_aligned_cleaned)]]
fut2_crop_cleaned <- fut2_crop[[names(worldclim_data_aligned_cleaned)]]

```


```{r Run the maxent models, include=FALSE, echo=FALSE}
#### create test and train data ####
# Now let us divide this file into 5 parts. 75% of the data will be used to train the model (to create the mode), and 25% will be used to test if it is a good model.
group <- kfold(points,5)
pres_train <- points[group!=1,]
pres_test <- points[group==1,]

#### modeling predictor contribution with maxent and create pseudo-absences ####
model_full <- maxent(predictors_cleaned, pres_train, removeDuplicates = TRUE)
model_bioclim <- maxent(worldclim_data_aligned_cleaned, pres_train, removeDuplicates = TRUE)
model_landuse <- maxent(landuse_rasterbrick_cleaned, pres_train, removeDuplicates = TRUE)

# create pseudo-absences to evaluate the model using AUC
predictorset <- c(predictors_cleaned, worldclim_data_aligned_cleaned, landuse_rasterbrick_cleaned)
models <- c(model_full, model_bioclim, model_landuse)
plot_titles <- c("Full model", "BioClim model", "Land Use model")

evalplots <- list()
AUC <- c()
for (i in c(1:3)) {
  backg = randomPoints(predictorset[[i]], n=10000, ext = study_area_ext, extf=1.25)
  colnames(backg) <- c("lon","lat")
  group=kfold(backg, 5)
  backg_train <- backg[group!=1,]
  backg_test <- backg[group==1,]
  e = evaluate(pres_test, backg_test, models[[i]], predictorset[[i]])
  # Extract TPR and FPR for ROC curve
  TPR <- e@TPR  # True Positive Rate
  FPR <- e@FPR  # False Positive Rate
  AUC <- round(e@auc, 3)
  # Store the ROC plot as a ggplot object in evalplots
  evalplots[[i]] <- ggplot() +
    geom_line(aes(x = FPR, y = TPR), color = "blue") +
    xlab("False Positive Rate") + 
    ylab("True Positive Rate") +
    labs(title = plot_titles[i],
         subtitle = paste("AUC =",AUC)) +
    theme_minimal()
}
```
## Predictor contributions for every model

 - *Full model*: This model includes all 19 BioClim variables all land use variables as predictors.
 - *BioClim model*: This model exclusively uses the BioClim variables to train the model.
 - *Land Use model*: This model exclusively uses the land use variables to train the model.
 
**BioClimatic variables**<br>
BIO1 = Annual Mean Temperature<br>
BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp))<br>
BIO3 = Isothermality (BIO2/BIO7) (×100)<br>
BIO4 = Temperature Seasonality (standard deviation ×100)<br>
BIO5 = Max Temperature of Warmest Month<br>
BIO6 = Min Temperature of Coldest Month<br>
BIO7 = Temperature Annual Range (BIO5-BIO6)<br>
BIO8 = Mean Temperature of Wettest Quarter<br>
BIO9 = Mean Temperature of Driest Quarter<br>
BIO10 = Mean Temperature of Warmest Quarter<br>
BIO11 = Mean Temperature of Coldest Quarter<br>
BIO12 = Annual Precipitation<br>
BIO13 = Precipitation of Wettest Month<br>
BIO14 = Precipitation of Driest Month<br>
BIO15 = Precipitation Seasonality (Coefficient of Variation)<br>
BIO16 = Precipitation of Wettest Quarter<br>
BIO17 = Precipitation of Driest Quarter<br>
BIO18 = Precipitation of Warmest Quarter<br>
BIO19 = Precipitation of Coldest Quarter

**Land use variables**<br>
Values are the fraction of a landcover class in each cell. The values are derived from the ESA WorldCover data set at 0.3-seconds resolution.<br>
Landcover variables:<br>
  - Trees<br>
  - Grassland<br> 
  - Shrubs<br>
  - Water<br>
  - Wetland<br>
  - Bare<br>
  - Built<br>
Following variables are also included:<br>
  - Population density<br>
  - Human footprint<br>
  - Elevation

more info on: https://github.com/rspatial/geodata

```{r Plotting predictor contributions, include=TRUE, echo=FALSE}
par(mfrow = c(3, 1), mar = c(2, 10, 1.5, 1))
plot(model_full, main = "Full model") 
plot(model_bioclim, main = "BioClim model")
plot(model_landuse, main = "Land Use model")
par(mfrow = c(1,1))
```

Predictors within every model have been purged for correlation if r > 0.75. When pairwise correlation is detected above this threshold, one of the variables within the pair is omitted before further processing.



## Model validation

In order to validate our models, we have used 75% of the data to train our MaxEnt (Maximum Entropy) model, while keep the remaining 25% aside for model validation.
The ROC curve is drawn by calculating the true positive rate (TPR) and false positive rate (FPR) at every possible threshold, then graphing TPR over FPR. A perfect model has a TPR of 1.0 and a FPR of 0.0. The larger the Area Under the Curve (AUC), the better a model performs.

```{r Plotting model accuracy, include=TRUE, echo=FALSE}
cowplot::plot_grid(plotlist = evalplots, ncol = 3)
```

```{r Maxent predictions, include=FALSE, echo=FALSE}
#### Create a map of the predicted distributions in the present and future ####
# PRESENT SDM for all predictors, bioclim only and landuse only
present_full <- predict(x = predictors_cleaned, object = model_full, ext = study_area_ext, progress='')
present_bioclim <- predict(x = worldclim_data_aligned_cleaned, object = model_bioclim, ext = study_area_ext, progress='')
present_landuse <- predict(x = landuse_rasterbrick_cleaned, object = model_landuse, ext = study_area_ext, progress='')

# FUTURE SDM for BioClim predictors
fut_pred <- predict(fut_crop_cleaned, model_bioclim, ext = study_area_ext, progress='', na.rm = TRUE)
fut2_pred <- predict(fut2_crop_cleaned, model_bioclim, ext = study_area_ext, progress='', na.rm = TRUE)

```


# MaxEnt SDM
## Present: 
```{r, plotting the present, include=TRUE, echo=FALSE}
plot(present_full)
title(bquote('MaxEnt full model ' * italic(.(params$Species))))
# shapefile of countries
data(outline, package = "ggmap2")
all_countries <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
plot(all_countries, add = TRUE, col = "transparent", border = "black")
# add occurrences with correct projection
plot(spdf, add = TRUE, pch = 16, cex = 0.5, col = rgb(red = 0, green = 0, blue = 1, alpha = 0.4))

plot(present_bioclim)
title(bquote('MaxEnt bioclim model ' * italic(.(params$Species))))
# shapefile of countries
data(outline, package = "ggmap2")
all_countries <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
plot(all_countries, add = TRUE, col = "transparent", border = "black")
# add occurrences with correct projection
plot(spdf, add = TRUE, pch = 16, cex = 0.5, col = rgb(red = 0, green = 0, blue = 1, alpha = 0.4))

plot(present_landuse)
title(bquote('MaxEnt landuse model ' * italic(.(params$Species))))
# shapefile of countries
data(outline, package = "ggmap2")
all_countries <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
plot(all_countries, add = TRUE, col = "transparent", border = "black")
# add occurrences with correct projection
plot(spdf, add = TRUE, pch = 16, cex = 0.5, col = rgb(red = 0, green = 0, blue = 1, alpha = 0.4))

```

## Future:
```{r, plotting the future, include=TRUE, echo=FALSE}
plot(fut_pred)
title(bquote('MaxEnt model ' * italic(.(params$Species)) * ' : 2041-2060'))
data(outline, package = "ggmap2")
all_countries <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
plot(all_countries, add = TRUE, col = "transparent", border = "black")
plot(spdf, add = TRUE, pch = 16, cex = 0.5, col = rgb(red = 0, green = 0, blue = 1, alpha = 0.4))

plot(fut2_pred)
title(bquote('MaxEnt model ' * italic(.(params$Species)) * ' : 2061-2080'))
data(outline, package = "ggmap2")
all_countries <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
plot(all_countries, add = TRUE, col = "transparent", border = "black")
plot(spdf, add = TRUE, pch = 16, cex = 0.5, col = rgb(red = 0, green = 0, blue = 1, alpha = 0.4))

```

